Supplement to *Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health for a Data Analysis Project

### 2023-05-06 3. Analysis Script/StatisticalAnalysis.qmd
Good Night, and Good Luck

```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #For plotting/data vis.
library(broom) #For cleaning up output from lm()
library(dotwhisker) #Visualization for regression outputs
library(here) #Setting paths/for data loading/saving
library(tidymodels) #Great package for models and other resources used for fitting
library(tidyr) #Helps with data wrangling
library(gtsummary) #To create tables 
library(dplyr)
library(vip)
library(performance) #Model evaluation
library(jtools)
library(broom)
library(broom.mixed) #Converts fitted to tiddy dfs
library(finalfit) #Needed for Time to Event Analysis 
library(forcats) #Needed for Time to Event Analysis 
library(survival) #Needed for Time to Event Analysis 
library(lubridate)
library(ggsurvfit)
library(gt)
library(webshot) #To save tables
library(knitr) #For tables
```

@@

CR_FD1%>%
  saveRDS(here("4. Results", "FD_Cox.rds"))
```
It looks like the

>>>>

```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #For plotting/data vis.
library(broom) #For cleaning up output from lm()
library(dotwhisker) #Visualization for regression outputs
library(here) #Setting paths/for data loading/saving
library(tidymodels) #Great package for models and other resources used for fitting
library(tidyr) #Helps with data wrangling
library(gtsummary) #To create tables 
library(dplyr)
library(vip)
library(performance) #Model evaluation
library(jtools)
library(broom)
library(broom.mixed) #Converts fitted to tiddy dfs
library(finalfit) #Needed for Time to Event Analysis 
library(forcats) #Needed for Time to Event Analysis 
library(survival) #Needed for Time to Event Analysis 
library(lubridate)
library(ggsurvfit)
library(gt)
library(knitr) #For tables
```

@@

CR_FD1%>%
  saveRDS(here("4. Results", "FD_Cox.rds"))
```
With our models run, let's see which one did the best. Because we have a small sample size, we will run an AICc.
```{r}

AIC_TAB_picnic<- compare_performance(lmfit_PT,lmfit_PT1,lmall_PT, lmfit_PT_glm, lmfit_PTglm1, lmfit_PTglm2, lmall_PTGLM, rank= TRUE, metrics="AICc")

print(AIC_TAB_picnic)

AIC_TAB_picnic %>%
  saveRDS(here("4. Results", "LMModelAICPD.rds"))

AIC_TAB_feeder<- compare_performance(lmfit_FD_glm, lmfit_FD, lmfit_FD2, lmall_FD, lmfit_FD_GLM, lmall_FDGLM_all1, rank= TRUE, metrics="AICc")

print(AIC_TAB_feeder)

AIC_TAB_feeder%>%
  saveRDS(here("4. Results", "ModelAICFD.rds"))
```

### 2023-05-06 5. Manuscript/Manuscript.qmd
Good Night, and Good Luck

-   What is the prevalence of *Salmonella* over time?

@@

# **Statistical analysis**

Statistical analysis was conducted on the Picnic Table Data and Feeder
Data, utilizing RStudio "Cherry Blossom" (2023-03-09). Analysis for both
datasets were evaluated utilizing several models including Linear
Regression (LR), Generalized Linear Model (GLM), and Cox Regression
(CR). Given that two datasets were utilized for this analysis, a
holistic approach, considering various limitations of both trials, such
as sample size, was taken into account when selecting a model. Because
the majority of data is binary in nature (e.g., presence v. absences)
and few variables exist for comparison, a LR was the first model
selected to run on both datasets, followed by a GLM, and CR. Several
predictors were selected to include in this the analysis including
'Salmonella_Positive\~Date' , 'Salmonella_Positive\~Sample_Type' (e.g.,
pooled v. persistent), and 'Salmonella_Positive\~.' with all predictors
for the Picnic Table Trial and 'Absence_0\_Presence_1\~Feeder_number',
'Absence_0\_Presence_1\~Feeder_number+Feeder_Type', and
'Absence_0\_Presence_1\~.' and all predictors for the Feeder Trial.

Both Linear Regression and Generalized Linear Models performed as
expected when model performance was checked utilizing the check_models()
function. Given the flexibility and model performance of the GLM, it was
selected as the model for inclusion in our analysis between predictors
(e.g., 'Date' \[PT\], 'Sample_Type' \[PD\], 'Feeder_Type' \[FD\]) and
outcomes (e.g., 'Salmonella_Positive' \[PT\], 'Absence_0\_Presence_1'
\[FD\]). To

While LR and GLM provide insight into

@@

**Picnic Table Data Exploration and Visualization**

@@

@@

**Bird Feeder Data Exploration and Visualization**

@@

## **Picnic Table Trial**

### **Persistence**

On Day 1 of the study, four fecal piles per table (one per quadrant, n =
12) were designated for tracking *Salmonella* persistence.
**Salmonella** was isolated from three of the 12 designated persistence
study piles on Day 1, allowing the persistence to be tracked on these
three piles. Persistence, expressed as the number of days that a
positive fecal pile on Day 1 remained positive, ranged from 1 to 4 days
with an average persistence of 3 days. The probability of an already
positive sample to remain positive the next day declined consistently
throughout the study period.

### **Prevalence**

Of 96 deposited feces, we obtained 27% positive. *Salmonella* persisted
for an average of 3 days, and maximum of 4 days on wooden picnic tables.
*Salmonella* was isolated from all tables, at least once. While there
appears to be a significant positive correlation between the number of
new fecal piles and the probability of yielding a positive pooled
sample. The probabilities of a positive sample varied throughout the
week, with day 4 having the lowest probability of positive *Salmonella*
isolation and day 7 having the most error.

## **Feeder Trial**

We confirmed successful inoculation of all feeders with *Salmonella*
(presence on day post inoculation \[DPI\] 0). Further, results indicate
that *Salmonella* may persist on some feeders up to DPI 5 and current
analysis indicates that *Salmonella* persisted on both coated and
uncoated plastic feeders at a higher rate than wood and wood coated.
Figure 2 depicts the percent of samples positive based on feeder type
throughout the study period. While plastic and plastic coated feeders
had an overall higher percent positivity, a spike in percent positivity
in wood feeders occurred on DPI 5 after a rain event. Further analyses
should be conducted to determine whether environmental conditions (e.g.,
rain event) had a significant influence on persistence.

## **Exploratory/Descriptive analysis**

## **Basic statistical analysis**

After the modules following Part 1, I am leaning towards running a GLM
for both given the binary responses for *Salmonella* persistence (e.g.,
presence v. absence). I need to meet with Dr. Hernandez to clear up some
questions I have on the picnic table data, specifically regarding the
introduction of new feces. I am curious if I should include other data
into the mix to make my analysis more robust. At the moment, the
variables of interest for both are limited (e.g., two for the picnic
table data, and one primary variable of interest for the feeder trial).

## **Full analysis**

# **Discussion**

## **Summary and Interpretation**

## **Strengths and Limitations**

## **Conclusions**

>>>>

-   What is the persistence of *Salmonella* over time?

@@

@@

## **Exploratory/Descriptive Analysis**

### **Picnic Table Data: Exploration and Visualization**

@@

Figure 1 depicts the the number of positive *Sallmonella* samples by
sample type, persistence or pooled. Pooled samples were collected from
every quadrant daily, by swabbing a sampling sponge along the picnic
table surface. Persistence piles, on the other hand were, selected based
on two criteria, quadrant and size of fecal pile. The persistence pile
was sampled everyday for 8 days. While the information presented in
Figure 1 depicts count data, it is a useful visual to convey numbers of
positive samples for each sample type based on day.

@@

Figure 2 depicts the the number of positive *Sallmonella* samples by
quadrant. While the information presented in Figure 2 depicts count
data, it is a useful visual to convey which quadrants may have been "hot
spots" for the presence of *Salmonella*.

**Persistence** On Day 1 of the study, four fecal piles per table (one
per quadrant, n = 12) were designated for tracking *Salmonella*
persistence. **Salmonella** was isolated from three of the 12 designated
persistence study piles on Day 1, allowing the persistence to be tracked
on these three piles. Persistence, expressed as the number of days that
a positive fecal pile on Day 1 remained positive, varied with an average
persistence of 3 days.

**Prevalence** Of 96 deposited feces, we obtained 27% positive (26/96).
*Salmonella* persisted for an average of 3 days, and maximum of 4 days
on wooden picnic tables. *Salmonella* was isolated from all tables, at
least once.

### **Bird Feeder Data Exploration and Visualization**

@@

Figure 3 depicts the the number of bird feeders that remained
*Salmonella* positive throughout the trial. Stacked bars in various
colors indicate the number of feeders (plastic, plastic + antimicrobial
coating, wood, and wood + antimicrobial coating) from which we isolated
*Salmonella*. We confirmed successful inoculation of all feeders with
Salmonella (presence on Day Post Inoculation (DPI) 0- 2/22/2022).
Further, results indicate that Salmonella may persist on some feeders up
to five DPI and current analysis indicates that Salmonella persisted on
both coated and uncoated feeders. With persistence occurring with both
plastic and plastic coated feeders.

# **Statistical Analysis**

Statistical analysis was conducted on the Picnic Table Data and Feeder
Data, utilizing RStudio "Cherry Blossom" (2023-03-09). Analysis for both
datasets were evaluated utilizing several models including Linear
Regression (LR) and Generalized Linear Model (GLM). Given that two
datasets were utilized for this analysis, a holistic approach,
considering various limitations of both trials, such as sample size, was
taken into account when selecting a model. Because the majority of data
is binary in nature (e.g., presence v. absence) and few variables exist
for comparison, a LR was the first model selected to run on both
datasets, followed by a GLM. Several predictors were selected to include
in this the analysis including 'Salmonella_Positive\~Date' ,
'Salmonella_Positive\~Sample_Type' (e.g., pooled v. persistent), and
'Salmonella_Positive\~.' with all predictors for the Picnic Table Trial
and 'Absence_0\_Presence_1\~Feeder_number',
'Absence_0\_Presence_1\~Feeder_number+Feeder_Type', and
'Absence_0\_Presence_1\~.' and all predictors for the Feeder Trial.

## **Regression Models**

### **Picnic Table Trial**

```{r}
#| label: Table 1
#| tbl-cap: "Small sample adjusted Akaike's Information Criteria (AICc) model weights and performance scores for linear and generalized linear models predicting Salmonella Persistance on Picnic Tables. "
#| echo: FALSE
LMModelAICPD = readRDS(here("4. Results", "LMModelAICPD.rds"))
knitr::kable(LMModelAICPD)
```

Of the models evaluated, the highest performing of all contained the
predictors original persistence pile \[Original_Persistence_Pile\],
table, and date and the outcome of *Salmonella* \[Salmonella_Positive\].
Given the AICc_wt along with performance score, a GLM was selected for
the Picnic Table Trial data, as depicted in Table 1. When evaluating a
model using AICc, we would ideally like to select a model with a low
AICc and a high performance score. Out of all of the models run, the
model labeled "lmfit_PTglm1" was selected. In this model, original
persistence pile along with table and date was associated with a higher
likelihood of Salmonella persistence.

### **Feeder Trial**

```{r}
#| label: Table 2
#| tbl-cap: "Small sample adjusted Akaike's Information Criteria (AICc) model weights and performance scores for linear and generalized linear models predicting Salmonella Persistance on Various Feeder Surfaces."
#| echo: FALSE
ModelAICFD= readRDS(here("4. Results", "ModelAICFD.rds"))
knitr::kable(ModelAICFD)
```

Of the models evaluated, the highest performing of all contained all
predictors. Given the AICc_wt along with performance score, a LM was
selected for the Feeder Trial data, as depicted in Table 2. When
evaluating a model using AICc, we would ideally like to select a model
with a low AICc and a high performance score, as mentioned above.
However, because this model can be considered "global" as it includes
all predictors, it is not the most ideal as it may overfit the data.
Despite this, the model labeled "lmall_FD" was selected. In this model,
feeder number, collection data, and feeder type was associated with a
higher likelihood of Salmonella persistence on feeders.

# **Discussion**
As cases of environmental salmonellosis increase throughout the United
States, environs that individuals come into contact with on a regular
basis need to be evaluated to fill knowledge gaps relating to microbial
transmission [@Hernandez2016]. With these trials we will research the
factors that influence *Salmonella* persistence, which will allow us to
better inform public health agencies and mitigate seasonal outbreaks in
humans and wildlife.

## **Summary and Interpretation**
Based on the modeling conducted above, there seems to be positive
effects of the various predictors on *Salmonella* absence/presence. The
Picnic Table Data showed that the original pile, date, and table all had
statistically significant effects on *Salmonella* presence on picnic
tables in South Florida. Persistence of *Salmonella* varied, however,
the maximum number of days it persisted was four
days (12 positive persistence piles/32 piles total).

Feeder data also indicate that *Salmonella* can persist on various surfaces for several days. *Salmonella* persisted for an average of ~2 days, (5 days max). Surprisingly, however, *Salmonella* persisted on plastic and plastic coated feeders longer than wooden and wooden coated feeders and had a prevalence of 15% on plastic coated feeders.

The results from both trials indicate that *Salmonella* can persist on various surfaces. As human and wildlife interactions increase due to urbanization, precautions should be taken to avoid potential pathogen transmission from the wildlife directly or through interactions with anthropogenic environs that may harbor wildlife excrement. While the data collected was mainly absence/presence, through data visualization, statistical analysis, and performance evaluations, gaps in knowledge surrounding *Salmonella* persistence outside of the poultry industry can be closed.

## **Strengths and Limitations**
While information on *Salmonella* persistence outside of the commercial poultry industry is relatively minimal, our work with the Picnic Table and Feeder Trials, provides some interesting insight into the persistence of this pathogen and possible sources of transmission during avian outbreaks (e.g., feeders). Because both experiments were trials, limited funding was allocated to each and a small sample size was a result of such. Small sample sizes limit data interpretation. Despite the small sample size, all models were analyzed with the model_check() function and most produced the expected results. However, some residuals in certain models were not normally distributed which can lead to over fitting.

Despite the limitations of these trials, steps were taken to close knowledge gaps. Trials, such as the ones conducted, can lay the groundwork for future studies that improve upon our sampling and collection methodology, as well as statistical analysis. Through the information garnered from these trials as well as future studies, more insight into *Salmonella* persistence and ways to mitigate it in our communities will inevitably be developed and enacted.

## **Conclusions**
Globally, salmonellosis outbreaks are common in aquatic and songbird populations and Salmonella continues to be a significant pathogen for public health. Notably, in 2021 a Salmonella Typhimurium outbreak in people was associated with the use and handling of bird feeders. Research is needed to understand the interaction between humans, birds, and their interface. Key to preventing outbreaks is understanding transmission dynamics, specifically, how long Salmonella persists on various materials. Persistence has been studied in food systems and commercial poultry operations—neither are relevant to free-living birds or people in contact with shared areas. Through further research and investment some of these pressing questions may be answered. Saving lives of various species globally.

### 2023-05-06 5. Manuscript/supplement/Supplementary_Material.qmd
Good Night, and Good Luck

---
title: "Supplement to Example Manuscript Template for a Data Analysis Project"

subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../dataanalysis_template_references.bib
csl: ../../vancouver.csl
---

# **Overview**
Overview of what I will include for readers needs to be included here.

# **Code and file information**

Here is where I will provide a step-by-step document so others will be able to follow along and easily reproduce my code. I will utilize some of the text I included in the main text/Manuscript.qmd and will expand upon it as I work through this project.

{{< pagebreak >}}

# **Additional Method Details**
{{< pagebreak >}}

# **Additional results**
{{< pagebreak >}}

# **Discussion**
{{< pagebreak >}}

# **References**

>>>>

---
title: "Supplement to *Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health for a Data Analysis Project"
author: Kimberly Perez
date: "`r Sys.Date()`"
output: html_document
---

# **Overview**
The information included on this .qmd file is code that was written but not utilized in final version of my manuscript. The majority of what you will find on this page comes directly from my "StatisticalAnalysis.qmd".

Loading in the packages we may need...
```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #For plotting/data vis.
library(broom) #For cleaning up output from lm()
library(dotwhisker) #Visualization for regression outputs
library(here) #Setting paths/for data loading/saving
library(tidymodels) #Great package for models and other resources used for fitting
library(tidyr) #Helps with data wrangling
library(gtsummary) #To create tables 
library(dplyr)
library(vip)
library(performance) #Model evaluation
library(jtools)
library(broom)
library(broom.mixed) #Converts fitted to tiddy dfs
library(finalfit) #Needed for Time to Event Analysis 
library(forcats) #Needed for Time to Event Analysis 
library(survival) #Needed for Time to Event Analysis 
library(lubridate)
library(ggsurvfit)
library(gt)
library(knitr) #For tables
```

We need the data from both trials, so let's load them in using here() 
```{r}
#Picnic Table Data
PT_Data<- readRDS(here::here("2. Clean Data","PTprocessed.rds"))

#Feeder Data 
FD_Data <-readRDS(here::here("2. Clean Data","FDprocessed.rds"))

glimpse(PT_Data)

glimpse (FD_Data)
```

```{r}
#### Model fit another way
# fit linear model using Salmonella_Pos as outcome, Table as predictor
lmfit3.0 <- lm(Salmonella_Positive~Table, PT_Data) 
lmfit3.0
plot(lmfit3.0)

# Here we place results from the above fit into a data frame via tidy function
lmtable1 <- broom::tidy(lmfit3.0)

#Let's check out the results
print(lmtable1)

#A different way 
# I will do the same as above, however, I am now adding persistence pile
lmfit2 <- lm(Salmonella_Positive~Original_Persistence_Pile+Table, PT_Data) 
plot(lmfit2)

# Convert to a data frame via tidy function
lmtable2 <- broom::tidy(lmfit2)

#Results
print(lmtable2)
```

This section is a continuation of what was produced above, however, we will now be utilizing the Feeder Trial Data

```{r}
#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

model_lmFD<- linear_reg()

lmfit_FD<- model_lmFD %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Absence_0_Presence_1~Feeder_number, FD_Data)

lmfit_FD #Let's view what we just did

glimpse(lmfit_FD)

tidy(lmfit_FD) #We can use the tidy() function to produce the output/ summary statistics

plot_summs(lmfit_FD)

#Checking model performance: FEEDER
check_model(lmfit_FD$fit)
```

**Time to Event Analysis: Picnic Table Trial**
```{r}

PT_Data_TTE<- #First let's format the dates 
  PT_Data %>% 
  mutate(
    Date= ymd(Date))

PT_Data_TTE$Date<- as.numeric(PT_Data_TTE$Date) #Let's make these numeric

PT_Data_TTE <- #I want to add a column displaying time between events (by day)
  PT_Data_TTE %>%
  mutate(
    os_day= as.duration(Date)/ddays(1))

PT_Data_TTE$os_day<- as.numeric(PT_Data_TTE$os_day) #Again let's make sure our new column is in the proper format

surv1<- Surv(PT_Data_TTE$Date,PT_Data_TTE$Salmonella_Positive)

#Let's visualize!
sfit1<- survfit(Surv(Date, Salmonella_Positive) ~  1, data=PT_Data_TTE)  %>% 
  ggsurvfit() + 
  labs(
    x = "Days",
    y = "Overall Salmonella Persistence") + #Let's add some labels
  add_confidence_interval() +
  add_risktable()

#I am not confident with this graphic so I will include this in my supplemental section
```

```{r}
glimpse(PT_Data)

PT_Data = PT_Data %>% 
 mutate(
  salmonella_crr= case_when(
   Salmonella_Positive ==  0~1,
    Salmonella_Positive == 1~1,
     TRUE ~ 1))

dependent_crr= "Surv(Date, salmonella_crr)"
explanatory= c("Table", "Quadrant", "Sample_Type")

PT_Data$Date<- as.numeric(PT_Data$Date)

PT_Data %>% finalfit(dependent_crr, explanatory)

PT_Data %>% 
   finalfit(dependent_crr, explanatory, add_dependent_label = FALSE) %>% 
    rename("Overall survival" = label) %>% 
    rename(" " = levels) %>% 
    rename("  " = all)

survfit(Surv(Date, Salmonella_Positive)~1, data=PT_Data)

PT_MedSurv<-PT_Data %>%
 filter(salmonella_crr==1) %>%
  summarize(median_surv=median(Date))

PT_MedSurv <- ((PT_MedSurv)/31556952000)*100

PT_Surv_Table<- gt(PT_MedSurv) %>%
   tab_header(
    title = "Median Survival of Salmonella",
    subtitle = ("01/02/2020 to 01/09/2020")) %>%
  cols_label(
    median_surv= "Median Salmonella Survival (days)")

PT_Surv_Table # persistence seems to be roughly 5 days
```

**Time to Event Analysis: Feeder Trial**
```{r}
FD_Data_TTE<- #First let's format the dates 
  FD_Data %>% 
  mutate(
    Date= ymd(Collection_Date))

FD_Data_TTE$Collection_Date<- as.numeric(FD_Data_TTE$Collection_Date) #Let's make these numeric

FD_Data_TTE <- #I want to add a column displaying time between events (by day)
  FD_Data_TTE %>%
  mutate(
    os_day= as.duration(Collection_Date)/ddays(1))

FD_Data_TTE$os_day<- as.numeric(FD_Data_TTE$os_day) #Again let's make sure our new column is in the proper format

surv1_FD<- Surv(FD_Data_TTE$Collection_Date,FD_Data_TTE$Absence_0_Presence_1)

#Let's visualize!
sfit1<- survfit(Surv(Collection_Date, Absence_0_Presence_1) ~  1, data=FD_Data_TTE)  %>% 
  ggsurvfit() + 
  labs(
    x = "Days",
    y = "Overall Salmonella Persistence") + #Let's add some labels
  add_confidence_interval() +
  add_risktable()

sfit1
```

```{r}
FD_Data = FD_Data %>% 
 mutate(
  salmonella_crr= case_when(
   Absence_0_Presence_1 ==  0~1,
    Absence_0_Presence_1 == 1~1,
     TRUE ~ 1))

dependent_crr= "Surv(Collection_Date, salmonella_crr)"
explanatory= c("Feeder_Type")

FD_Data$Collection_Date<- as.numeric(FD_Data$Collection_Date)

FD_Data %>% finalfit(dependent_crr, explanatory)

FD_Data %>% 
   finalfit(dependent_crr, explanatory, add_dependent_label = FALSE) %>% 
    rename("Overall survival" = label) %>% 
    rename(" " = levels) %>% 
    rename("  " = all)

survfit(Surv(Collection_Date, Absence_0_Presence_1)~1, data=FD_Data)

FD_MedSurv<-FD_Data %>%
 filter(salmonella_crr==1) %>%
  summarize(median_surv=median(Collection_Date))

FD_MedSurv <- ((FD_MedSurv)/604800)*100 #Persistence seems to be roughly 3 days for the FD

FD_Surv_Table<- gt(FD_MedSurv) %>%
   tab_header(
    title = "Median Survival of Salmonella",
    subtitle = ("02/22/2022 to 03/14/2022")) %>%
  cols_label(
    median_surv= "Median Salmonella Survival (days)")

FD_Surv_Table # seems to be roughly 3 days
```

**Cox regression: Picnic Table Trial**

```{r}
PT_Data$Date<- as.numeric(PT_Data$Date)

coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data)

CR_PT1<-coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data) %>%
  tbl_regression(exp=TRUE)

CR_PT1
```

**Cox regression: Feeder Trial**

```{r}
FD_Data$Collection_Date<- as.numeric(FD_Data$Collection_Date)

coxph(Surv(Collection_Date, Absence_0_Presence_1) ~ Feeder_Type, FD_Data)

CR_FD1<- coxph(Surv(Collection_Date, Absence_0_Presence_1) ~ Feeder_Type, FD_Data) %>%
  tbl_regression(exp=TRUE)

CR_FD1

CR_FD1%>%
  saveRDS(here("4. Results", "FD_Cox.rds"))
```
